---
title: "Ancestry based on genotype likelihoods"
output: github_document
---

We start by loading the data and identifying structure of the files:

```{r}
### Load data
pop <- read.table('pop.info', header=F)
colnames(pop) <- c('population', 'sample')
qopt <- read.table('assign3.qopt', header=F)

fopt <- read.table('assign3.fopt.gz', header=F)

input <- read.table('input.gz', header=T)

prob_geno <- input[,-c(1:3)]
#################### Identify columns ################################

# Identify populations
print(unique(pop$population))

# Identify columns
by(qopt, pop$population, summary)
```

We see that the first column corresponds to african, second to european and third to chinese.

```{r}
# Rename columns according to findings
colnames(qopt) <- c("African", "European", "Chinese")
```

Let's take a look at the average chinese ancestry in all individuals from the ASW population:

```{r}
# Compute average chinese ancestry
mean(qopt[which(pop$population == "ASW"),"Chinese"])
```


We are interested in testing the hypothesis that the chinese ancestry in the african american
individuals is due to statistical noise. For this test we use the following hypothesis:

```{r}
# Find admixture proportions for first ASW individual
qopt_null <- qopt
qopt_null[which(pop$population == "ASW"),"Chinese"] <- 0
qopt_null[which(pop$population == "ASW"),"European"] <- 
1 - qopt[which(pop$population == "ASW"),"African"]



print(c("Population admixture under alternative hypothesis for the first individual: ", 
        qopt[which(pop$population == "ASW")[1],]))
print(c("Population admixture under null hypothesis for the first individual: ",
        qopt_null[which(pop$population == "ASW")[1],]))
```

We set up the likelihood model to test our hypothesis:

```{r}
# Function to compute h
compute_PGh <- function(fopt, qopt, index){
hj <- as.matrix(fopt) %*% t(qopt[index,])
PGh <- cbind((1-hj)**2, 2*hj*(1-hj), (hj)**2)
return(PGh)
}

# Function to compute log likelihood
loglike <- function(fopt, qopt, prob_geno, ind, pop_info){
index <- which(pop_info$sample == ind)
PGh <- compute_PGh(fopt, qopt, index)
prob_geno_ind <- prob_geno[,(index*3-2):(index*3)]
loglike <- sum(log(apply(prob_geno_ind*PGh, 1, sum)))
return(loglike)
}
```

Now we can apply the likelihood model for the first individual and perform the 
likelihood ratio test to test our hypothesis:

```{r}
# Compute logLikelihood for first ASW individual
ind <- "NA19818"
# For each model
logLikeAlt <- loglike(fopt, qopt, prob_geno, ind, pop)
logLikeNull <- loglike(fopt, qopt_null, prob_geno, ind, pop)
# Compute likelihood ratio test
LR <- -2 *(logLikeNull - logLikeAlt)
pvalue <- 1 - pchisq(LR,df=1)
# Print results
print(rbind(ind, LR, pvalue))
```


Now let us do this for all african american individuals to find the number of 
individuals with significant amount of chinese ancestry:

```{r}
# Now compute log likelihood for all ASW individuals
asw_df <- data.frame(row.names = c('LR', 'pvalue'))
i <- 1
for (index in which(pop$population == 'ASW')){
ind <- pop$sample[index]
logLikeAlt <- loglike(fopt, qopt, prob_geno, ind, pop)
logLikeNull <- loglike(fopt, qopt_null, prob_geno, ind, pop)

LR <- -2 *(logLikeNull - logLikeAlt)
pvalue <- 1 - pchisq(LR,df=1)

asw_df <- cbind(asw_df, rbind(LR, pvalue))
colnames(asw_df)[i] <- ind
i <- i + 1 
}
print(asw_df)

sum(asw_df["pvalue",]<0.05)
```

